Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-18 mpaper/paper.tex
Sage plots of q(u) and its derivatives

\section{Introduction}
% problem description/statement

@@

\section{Background} \label{sec:background}
% motivation, relevance (whatever that means)
% comprehensive literature survey

@@

\subsection{Theoretical Justification}
% TODO: in this section, 'constant' means 'constant w.r.t. u and r'

@@

The typical way to optimise a quantity (the ELBO, in this case) involves
computing its gradient. Unfortunately, some of the terms in $\mathcal{L}$ are
still left as expected values. The goal of this section is to show how Theorem
\ref{thm:lebesgue} can be applied to our model in order to derive the gradient
anyway\footnote{This technique is inspired by black box VI
  \cite{DBLP:conf/aistats/RanganathGB14}, but takes a more detailed look at the
  problem and requires significantly more work to prove correctness.}. After
showing that the theorem applies to our situation, given any term expressed as
an expected value $\mathbb{E}[f(\mathbf{u}, \mathbf{r})]$, we can estimate its
gradient as
\[
  \nabla\mathbb{E}[f(\mathbf{u}, \mathbf{r})] = \mathbb{E}[\nabla f(\mathbf{u},
  \mathbf{r})] \approx \frac{1}{S} \sum_{s=1}^S \nabla f(\mathbf{u}_s,
  \mathbf{r}_s),
\]
which can be computed by drawing $S$ Monte Carlo samples $(\mathbf{u}_s,
\mathbf{r}_s) \sim \approximation$.

@@

Before that, however, we define a few extra variables in order to simplify
expressions of derivatives:
\begin{align*}
  \mathbf{U} &= (\mathbf{u} - \bm\mu)(\mathbf{u} - \bm\mu)^\intercal, \\
  \mathbf{S} &= \Kru^\intercal\Kuu^{-1}, \\
  \bm\Gamma &= \Krr - \mathbf{S}\Kru, \\
  \mathbf{R} &= \mathbf{S}\frac{\partial \Kru}{\partial \lambda_i} - \frac{\partial \Krr}{\partial \lambda_i} + \left( \frac{\partial \Kru^\intercal}{\partial \lambda_i} - \mathbf{S}\frac{\partial \Kuu}{\partial \lambda_i} \right) \Kuu^{-1}\Kru, \\
  Q &= (\mathbf{u} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{u} - \bm\mu).
\end{align*}

\begin{restatable}[Derivatives of PDFs]{lemma}{derivatives} \label{lemma:derivatives}
  \leavevmode
  \begin{enumerate}
  \item $\frac{\partial q(\mathbf{u})}{\partial \bm\mu} =
    \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} + \bm\Sigma^{-\intercal})(\mathbf{u}
    - \bm\mu)$.
  \item
    \begin{enumerate}
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} =
      \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
      - \bm\Sigma^{-\intercal})$.
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} =
      q(\mathbf{u})(\bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} -
      |\bm\Sigma|^{-1}\adj(\bm\Sigma))\mathbf{B}$.
    \end{enumerate}
  \item For $i = 0, \dots, d$,
    \begin{enumerate}
    \item
      \begin{align*}
        \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &= \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1} \tr(\mathbf{R} \adj(\bm\Gamma)) \\
                                                                          &- (\mathbf{r} - \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su})).
      \end{align*}
    \item For any covariance matrix $\mathbf{K}$,
      \[
        \frac{\partial \mathbf{K}}{\partial \lambda_i} =
        \begin{cases}
          \frac{1}{\lambda_i}\mathbf{K} & \text{if } i = 0, \\
          \mathbf{L} & \text{otherwise,}
        \end{cases}
      \]
      where
      \[
        L_{j,k} = k(\mathbf{x}_j, \mathbf{x}_k) \left( -\frac{1}{2}(x_{j,i} -
          x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2 \right).
      \]
    \end{enumerate}
  \end{enumerate}
\end{restatable}

@@

% TODO: a section about the algorithm

@@

@@

% How well does the variance prediction work?

>>>>

\section{Introduction}
% problem description/statement
% TODO: rewrite and make it good

@@

\section{Background} \label{sec:background}
% motivation, relevance (whatever that means)
% comprehensive literature survey
% TODO: need to add a literature survey here

@@

\subsection{Theoretical Justification}

@@

The typical way to optimise a quantity (the ELBO, in this case) involves
computing its gradient. Unfortunately, the term $\mathbb{E}[v]$ in $\mathcal{L}$
complicates the situation. The goal of this section is to show how Theorem
\ref{thm:lebesgue} can be applied to our model in order to derive the gradient
anyway\footnote{This technique is inspired by black box VI
  \cite{DBLP:conf/aistats/RanganathGB14}, but takes a more detailed look at the
  problem and requires significantly more work to prove correctness.}. After
showing that the theorem applies to our situation, we can express
$\nabla\mathbb{E}[v]$ as
\begin{align*}
  \nabla\mathbb{E}[v] &= \nabla \iint q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})\dx \\
                      &= \iint \nabla[v q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})]\dx \\
                      &= \iint \frac{\nabla[v q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})]}{q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})} q(\mathbf{r} \mid \mathbf{u}) q(\mathbf{u})\dx \\
                      &\approx \frac{1}{S} \sum_{s=1}^S \frac{\nabla[v q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})]}{q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})},
\end{align*}
which can be computed by drawing $S$ Monte Carlo samples $(\mathbf{u}_s,
\mathbf{r}_s) \sim \approximation$.

@@

Before that, however, we define a few extra variables in order to simplify
expressions of derivatives:
\begin{align*}
  \mathbf{U} &= (\mathbf{u} - \bm\mu)(\mathbf{u} - \bm\mu)^\intercal, \\
  \mathbf{S} &= \Kru^\intercal\Kuu^{-1}, \\
  \bm\Gamma &= \Krr - \mathbf{S}\Kru, \\
  \mathbf{R} &= \mathbf{S}\frac{\partial \Kru}{\partial \lambda_i} - \frac{\partial \Krr}{\partial \lambda_i} + \left( \frac{\partial \Kru^\intercal}{\partial \lambda_i} - \mathbf{S}\frac{\partial \Kuu}{\partial \lambda_i} \right) \Kuu^{-1}\Kru, \\
  Q &= (\mathbf{u} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{u} - \bm\mu).
\end{align*}
Also note that throughout this section the word `constant' means `constant with
respect to $\mathbf{u}$ and $\mathbf{r}$'.

\begin{restatable}[Derivatives of PDFs]{lemma}{derivatives} \label{lemma:derivatives}
  \leavevmode
  \begin{enumerate}
  \item $\frac{\partial q(\mathbf{u})}{\partial \bm\mu} =
    \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} + \bm\Sigma^{-\intercal})(\mathbf{u}
    - \bm\mu)$.
  \item
    \begin{enumerate}
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} =
      \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} -
      \bm\Sigma^{-1})$.
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} =
      q(\mathbf{u})(\bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} -
      \bm\Sigma^{-1})\mathbf{B}$.
    \end{enumerate}
  \item For $i = 0, \dots, d$,
    \begin{enumerate}
    \item
      \begin{align*}
        \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &= \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1} \tr(\mathbf{R} \adj(\bm\Gamma)) \\
                                                                          &- (\mathbf{r} - \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su})).
      \end{align*}
    \item For any covariance matrix $\mathbf{K}$,
      \[
        \frac{\partial \mathbf{K}}{\partial \lambda_i} =
        \begin{cases}
          \frac{1}{\lambda_i}\mathbf{K} & \text{if } i = 0, \\
          \mathbf{L} & \text{otherwise,}
        \end{cases}
      \]
      where
      \[
        L_{j,k} = k(\mathbf{x}_j, \mathbf{x}_k) \left( -\frac{1}{2}(x_{j,i} -
          x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2 \right).
      \]
    \end{enumerate}
  \end{enumerate}
\end{restatable}

@@

% Maybe?
%\subsection{Implementation}

@@

How well does the variance prediction work? It doesn't, but why?
TODO: Sage should consider 3x3 matrices
TODO: 4 plots---two my MATLAB, two my Sage; two for B[1, 1], two for B[2, 1].
The MATLAB plots show E[v] and dE[v]/dB, the Sage plots show q(u) and
dq(u)/dB. Function in blue, derivative in red, over a range of values.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{sage1}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{sage2}
\end{figure}

@@
